---
title: "Intro to ggplot2"
author: "Marian L. Schmidt"
date: "March 9, 2015"
output: html_document
---

###What is ggplot2?
+ An implementation of the "Grammar of Graphics" by Leland Wilkson  
+ Written by Hadley Wickham  
+ Built on the grid graphics  
+ 3rd mode of plotting:  Base, lattice and ggplot  

###Anatomy of a ggplot:  
+ data  
+ faceting
+ aesthetic mapping  
+ geometric object  
+ statistical transformations  
+ scales  
+ coordinate system  
+ position adjustment


###Data:  
+ Always in a data.frame form! 

###Many people have published on how to use ggplot  
+ [R Cookbook](http://www.cookbook-r.com/Graphs/)  
+ [Beautiful Plotting with ggplot](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)  
+ [ggplot Handout from Hadley Wickham](http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf)  


###Today:  
+ **Barplots with ggplot:** very simple!  Can make it look like base R
+ **Scatterplot:**  Legends and coloring of points with ggplots "scales"  
+ **Heatmap:**  Faceting and Gradient Coloring  
+ **Multiplotting with ggplot:**  gridExtra


###Load packages
```{r load packages, echo= TRUE, message = FALSE}
library(ggplot2)
library(gridExtra) # Allows us to plot multiple plots together in one graph
```



```{r relative abundance, message=FALSE, echo = TRUE, warning = FALSE, results ='hide', fig.height=9, fig.align='center', fig.width = 8}
#Load Data
relabund_data <- read.table("relabund_data", sep = "\t", header = T)

#Order by the SeqAbundance
relabund_data <- relabund_data[order(relabund_data$PercentPhy),]
relabund_data$Phylum <- factor(relabund_data$Phylum, levels = relabund_data$Phylum[order(relabund_data$PercentPhy)])
relabund_data <- subset(relabund_data, Phylum != "unclassified" & Phylum != "Synergistetes" & Phylum != "RsaHF231" & Phylum != "FGL7S" & Phylum != "JTB23")
phys <- as.character(relabund_data$Phylum)

#Only plot Phylum that are more than 1%
abund <- subset(relabund_data,PercentPhy > 0.01)

#Let's look at the data
head(relabund_data)

#Making the Relative Abundance plot 
abund_plot <- ggplot(abund, aes(x = Phylum, y=PercentPhy))  +
  geom_bar(stat="identity", position=position_dodge(),  fill = "magenta4", colour = "black") +
    # Here stat="identity" is plotting a value, Default is stat="bin" which is count data 
  theme_bw() + ggtitle("Phyla Above 1% in All Samples") +
  xlab("Phylum") + ylab("Mean Relative Abundance (%)") +
  geom_errorbar(aes(ymin = PercentPhy -se, ymax = PercentPhy +se), width = 0.25) + coord_flip() +
    ### Here is where we set all of axis and plot themes for pretty plotting
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x = element_text(angle=0, colour = "black", size=14),
        axis.text.y = element_text(colour = "black", size=14),
        axis.title.y = element_text(face="bold", size=16),
        plot.title = element_text(face="bold", size = 20),
        legend.title = element_text(size=12, face="bold"),
        legend.text = element_text(size = 12),
        legend.position="none"); abund_plot
 
# We can overwrite old preferences 
abund_plot + theme_grey()

# Same asethetic as base R
abund_plot + theme_classic()

```




```{r nmds, message=FALSE, echo = TRUE, warning = FALSE, results ='hide', fig.height=8, fig.align='center', fig.width = 9}
#Load Data
nmds_data <- read.delim("NMDS_data", header = TRUE)

#Let's look at the data
head(nmds_data)
  
### TROPHIC STATE:  Combine Mesotrophic and Eutrophic
nmds_data$trophicstate[nmds_data$trophicstate == "Mesotrophic"] <- "Eutrophic"

ggplot(nmds_data, aes(x = MDS1,y = MDS2, color = trophicstate)) +
  xlab("NMDS1") + ylab("NMDS2") + scale_y_continuous(breaks=seq(-1, 1, 0.25)) + # Ticks from 0-10, every .25
  scale_x_continuous(breaks=seq(-1, 15, 0.5)) +
  annotate("text", label = "A", x = -0.7, y = 0.4, size = 14, colour = "black", face = "bold") +
  annotate("text", label = "Stress = 0.17", x = 0.60, y = 0.4, size = 6, colour = "black") +
  geom_point(size= 6, alpha=0.9) + theme_bw() +
  scale_color_manual(name = "Trophic State", labels = c("Eutrophic + Mesotrophic", "Oligotrophic", "Mixed"), values = c("deeppink", "turquoise3", "goldenrod")) +
  theme(axis.text.x = element_text(colour="black", vjust=0.5, size=14), 
        axis.text.y = element_text(colour="black", vjust=0.5, size=14),
        axis.title.x = element_text(face="bold", size=16),
        axis.title.y = element_text(face="bold", size=16),
        legend.title = element_text(size=12, face="bold"),
        legend.text = element_text(size = 12),
        legend.justification=c(1,0), legend.position=c(1.01, -0.01))

## NMDS colored by quadrant 
ggplot(nmds_data, aes(x = MDS1,y = MDS2, color = quadrant, shape = quadrant)) +
  xlab("NMDS1") + ylab("NMDS2") + scale_y_continuous(breaks=seq(-1, 1, 0.25)) + # Ticks from 0-10, every .25
  scale_x_continuous(breaks=seq(-1, 15, 0.5)) + # Ticks from 0-10, every .25
  annotate("text", label = "B", x = -0.7, y = 0.4, size = 14, colour = "black", face = "bold") +
  annotate("text", label = "Stress = 0.17", x = 0.60, y = 0.4, size = 6, colour = "black") +
  geom_point(size= 6, alpha=0.9) + theme_bw() +
  scale_color_manual(name = "Habitat", breaks=c("Free Top", "Free Mixed",  "Free Bottom", "Particle Top", "Particle Mixed", "Particle Bottom"),
                     labels = c("Free-Living Epilimnion", "Free-Living Mixed",  "Free-Living Hypolimnion", "Particle-Associated Epilimnion", "Particle-Associated Mixed", "Particle-Associated Hypolimnion"), 
                     values = c("magenta4", "dodgerblue", "forestgreen", "magenta", "dodgerblue", "chartreuse3")) +
  scale_shape_manual(name = "Habitat", breaks = c("Free Top", "Free Mixed",  "Free Bottom", "Particle Top", "Particle Mixed", "Particle Bottom"),
                     labels = c("Free-Living Epilimnion", "Free-Living Mixed",  "Free-Living Hypolimnion", "Particle-Associated Epilimnion", "Particle-Associated Mixed", "Particle-Associated Hypolimnion"), 
                     values = c(19, 19, 19, 17, 17, 17)) +
  theme(axis.text.x = element_text(colour="black", vjust=0.5, size=14), 
        axis.text.y = element_text(colour="black", vjust=0.5, size=14),
        axis.title.x = element_text(face="bold", size=16),
        axis.title.y = element_text(face="bold", size=16),
        legend.title = element_text(size=12, face="bold"),
        legend.text = element_text(size = 12),
        legend.justification=c(1,0), legend.position=c(1.01, -0.01)) 
```





### Faceting with ggplot 
```{r heatmaps, fig.width=11, fig.height = 14, fig.align='center', echo = TRUE, warning = FALSE, message = FALSE, results ='hide'}
#Load Data
heatmap_data <- read.table("heatmap_data", sep = "\t", header = T)

#Look at data
head(heatmap_data)

#Creating Facet Labels
facet_labels <- function(var, value){
    value <- as.character(value)
    if (var=="Comparison") { 
        value[value=="PA vs. FL"] <- "Particle-Associated \n vs. Free-Living"
        value[value=="Top vs. Bottom"]   <- "Hypolimnion \n vs. Epilimnion"
        value[value=="Prod vs. Unprod"]   <- "Productive \n vs. Unproductive"
    }
    return(value)
}



heat1 <- ggplot(heatmap_data, aes(x = Habitat, y = Phylum)) + geom_tile(aes(fill = Transformed)) + 
  scale_fill_gradient2(name = "Odds-Ratio", mid = "gray", low = "darkorange", high = "blue4",  na.value = "white") + #scale_y_reverse() + 
  theme_bw(base_size = 12) + scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) + ylab(NULL) + 
  #geom_text(aes(fill = splif2$Transformed, label = splif2$Transformed, size = 8)) +
  scale_y_discrete(limits=phys) + xlab("Habitat") + ylab("Phylum") + 
  facet_grid(. ~ Comparison, scales = "free", space = "free", labeller=facet_labels) + 
  theme(axis.text.x = element_text(colour="black", size=14, angle = 60, hjust = 1, vjust = 1), 
        axis.text.y = element_text(colour="black", vjust=0.5, size=14),
        axis.title.x = element_blank(), #text(face="bold", size=16),
        legend.title = element_text(face="bold", size=12),
        legend.text = element_text(size = 12),
        legend.position = "left",
        axis.title.y = element_text(face="bold", size=16),
        plot.margin = unit(c(0.5, 1, 0.5, 0.5), "cm"),
        strip.text.x = element_text(size=16, face = "bold", colour = "white"),
        strip.background = element_rect(colour="black", fill = "black")); heat1
```


### Multiplotting with ggplot
```{r putting together abundance and heatmap, fig.width=14, fig.height = 14, fig.align='center', echo = TRUE, warning = FALSE, message = FALSE, results ='hide'}
relabun_plot<- ggplot(relabund_data, aes(y=PercentPhy , x=Phylum)) + #coord_cartesian(xlim = c(0, 30)) + 
  geom_bar(stat="identity", position=position_dodge(),fill = "magenta4", colour = "black") +
  ylab("Mean Relative \n Abundance (%)") + coord_flip() + theme_bw() + 
  geom_errorbar(aes(ymin = PercentPhy -se, ymax = PercentPhy +se), width = 0.25, color = "black") + 
  scale_y_continuous(expand= c(0,0), limits = c(0,25)) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x = element_text(angle=0, colour = "black", size=14),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        strip.background = element_rect(colour="black", fill = "black"),
        plot.margin = unit(c(1.9, 2, 4.4, -1.25), "cm"), #top, right, bottom, left  #bottom = 3.4
        #panel.grid.minor=element_blank(), #panel.grid.major=element_blank(),
        legend.position="none")


#http://stackoverflow.com/questions/20817094/how-to-control-width-of-multiple-plots-in-ggplot2
grid.newpage()
pushViewport(viewport(layout=grid.layout(1,2,width=c(0.8,0.2))))
print(heat1, vp=viewport(layout.pos.row=1,layout.pos.col=1))
print(relabun_plot, vp=viewport(layout.pos.row=1,layout.pos.col=2))
```

